The tree line ecotone punctuates the elevational turnover of plants, pollinators, and their interactions 

### 2021-03-15 ms_1/ms_1.Rmd
refined GAM analysis

```{r fig1, echo = FALSE,  fig.cap = paste("Study sites in the Berchtesgadener Alps plotted over Google Earth imagery. Sites above the tree line are depecited in red and sites below the tree line in yellow.")}
knitr::include_graphics("../sites2_2021-03-11_17-40-25.png")
```

@@

### Distribution of bumble bees and floral resources in elevation and time

We analyzed bumble bee abundance and floral abundance using separate hierarchical generalized additive models (HGAM) [@Wood2011-pg; @Pedersen2019-kf], with abundance modeled as a response to the interaction (tensor product) of elevation and day-of-year. Bumble bee species was included in each model as both a smoothing factor and a random intercept term, resulting in centered 2-dimensional abundance smooths for each bumble bee species. We also included year and site as random intercept effects to account for annual differences in mean abundance and repeated measurements within sites.

@@

For our analyses of species and interaction β-diversity, we opted to focus on elevational patterns by pooling samples within sites, since the sparse observations of individual sampling events tended to inflate β-diversity and its variance to a degree that impaired model fitting and interpretation. After pooling, we calculated the species and interaction β-diversity between all pairs of sites and partitioned total interaction β-diversity into its components of species turnover (changes in species composition) and interaction rewiring (identical species interacting in different ways) [@Novotny2009-my; @Poisot2012-fk]. As an initial analysis, we plotted the relationship between site-wise elevation difference and each metric of β-diversity and verified the significance of the relationship using logistic matrix regression [@Goslee2009-ln], with the difference in number of sampling dates between sites as a covariate to control for potential confounding effects of sampling frequency.

@@

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. GAM analyses were performed with packages `mgcv` [@Wood_2017] and `mgcViz` [@Fasiolo2018aa], and the models we fit correspond to the “type I” model form described in @Pedersen2019-kf. PCoA and species β-diversity calculation were performed with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. Matrix regression was performed with the package `ecodist` [@Goslee2007-aa]. GDM analysis of β-diversity was performed with the package `gdm` [@Fitzpatrick-2021aa]. Annotated `R` code is available in the Supplementary Material.

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species, 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. Five species --- *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris/lucorum*, and *B. wurflenii* --- accounted for the bulk of overall bumble bee abundance and were present over the entire elevation range of our sites **(Figure 1)**. *B. hortorum*, *B. jonellus*, and *B. psithyrus* (species-group) were widespread but at lower abundance. *B. monticola* occurred at moderate abundance but was rare or absent at sites below 1000 m.s.l. The remaining species were relatively rare in our study system. *B. mendax*, *B. mucidus*, and *B. pyrenaeus* occurred primarily (peak abundance > 1) above the 1500 m.s.l., consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. *B. gerstaeckeri*, *B. hypnorum*, *B. lapidarius* were sparsely distributed across low- and mid-elevation sites. *B. humilis* was recorded only three times and only in one year, so we omitted it from all analyses.

```{r fig2, echo = FALSE,  fig.cap = paste("Peak abundances and elevational distributions of bumble bee species. Sampled sites along the elevation gradient are depicted as circles, the diameter of which is scaled to the peak abundance (pooled across study years) of the given bumble bee species. Abundances of zero are not plotted. Lines span between the highest and lowest occurrence of each species, highlighting its observed elevation range. Species are plotted in order of increasing median elevation.")}
knitr::include_graphics("../output/bb_range_ms1.png")
```

@@

## Distribution of bumble bees and floral resources in elevation and time

Cluster **(Figure 2A)** and ordination **(Figure 2B)** analyses highlighted the strong differences in site affinity between the high-elevation species *B. mucidus*, *B. monticola*, *B. mendax*, and *B. pyrenaeus* and the rest of the bumble bee community. PCoA captured 59% of total variation in site affinity.

Elevation-time GAM smooths **(Figure 2C)** of bumble bee abundance differed significantly from a flat surface (p < 0.00001) for all species except *B. hypnorum* (p = 0.8). Overall, the model explained 57.6% of total deviance (adjusted R^2^ = 0.346). The distinction between *B. mucidus*, *B. monticola*, *B. mendax*, and *B. pyrenaeus* and the rest of the community was again clear; all 4 high-elevation species exhibited peak abundance at the upper extreme of our elevation gradient, and only *B. monticola* was found at high abundance below 1500 m. *B. monticola* also peaked in abundance earlier in the season than the other high-elevation species.

*B. gerstaeckeri* and *B. pratorum* exhibited well-defined mid-elevation abundance peaks, with that of *B. gerstaeckeri* occurring very late in the season. The apparent mid-elevation abundance peak of *B. hypnorum* was not significant and should not be interpreted strongly.

*B. pascuorum* and *B. psithyrus* had well-defined low-elevation peaks, with that of *B. psithyrus* occurring early in the season, during the colony-founding --- or, in the case of *B. psithryus*, colony-usurping --- phase of the bumble bee life cycle.

The remaining species exhibited relatively diffuse patterns of abundance. Of these, *B. soroeensis* was exceptional in that it exhibited peak abundance at the highest sites, but it was also found at relatively high abundance across the whole elevation gradient, peaking at mid-season. *B. hortorum* and *B. lapidarius* peaked at low elevation. *B. jonellus* exhibited little elevational variation in abundance until declining above 1500 m, but its abundance was strongly patterned in time, peaking early in the season. *B. terrestris/lucorum* was generally abundant across the whole elevation gradient and throughout the whole season, but it showed a weakly trimodal pattern in elevation, with abundance peaks at low, middle, and high sites. Its strongest abundance peak occurred at low elevation, early in the season during colony-founding. *B. wurflenii* was broadly abundant above 1000 m but was rarer at lower sites.

Elevation-time GAM smooths of floral abundance differed significantly from a flat surface for all species (p << 0.00001) and explained 64.8% of total deviance (adjusted R^2^ = 0.425). Floral resource availability exhibited similar patterns for most bumble bee species, with peak floral resources occurring below 1500 m and during mid-season **(Figure 2)**. There was a noticeable temporal lag in floral resource availability associated with increasing elevation, consistent with expected temperature constraints on plant phenology. The only species to exhibit marked anomalies in floral resource availability was *B. gerstaeckeri*, which specializes on the genus *Aconitum* [@Ponchau2006-af] but was also observed (presumably collecting nectar) on *Gentiana*, *Salvia*, *Rhinanthus*, *Cirsium*, and *Clinopodium.* Its floral resources were concentrated in two discrete peaks, the first at low elevation and mid-season, and the second at high elevation and late-season.

```{r fig3, echo = FALSE,  fig.cap = paste("Abundance of bumble bee species and their floral resources through elevation and time. For each bumble bee species, surface plots are arranged in two columns labeled BB and FL. Surface plots in the BB columns show the abundance of the bumble bee species, while plots in the FL columns show the abundance of floral resources for the corresponding bumble bee species. The color ramp of each smooth has been scaled to equal color range, thereby obscuring the absolute magnitude of effect. To avoid cluttering the figure, we have opted not to include a unique color ramp key for each plot, since the focus of our inference is on relative rather than absolute patterns. Surface plots with original color ramp scales are available in Appendix X.")}
knitr::include_graphics("../output/gam_BB_FL4.png")
```

@@

```{r fig4, echo = FALSE,  fig.cap = paste("Species-level (A) and interaction-level (B, C) plotted against elevational difference. Each point represents the β-diversity (or β-diversity partition) between a pair of sites, and lines represent the overall relationship between β-diversity and elevation difference using binomial regression smooths. Standard errors are not plotted because they would be misleading due to the non-independence inherent to distance matrix regression, but all regressions were significant (p < 0.01). Interaction β-diversity (B, C) is partitioned using Poisot et al.’s (2012) notation: WN = unpartitioned β-diversity, ST = β-diversity due to species turnover, OS =  β-diversity due to interaction rewiring, ST.h =  β-diversity due to species turnover in the higher trophic level (bumble bees), ST.l = β-diversity due to species turnover in the lower trophic level (plants), and ST.lh =  β-diversity due to joint species turnover in higher and lower trophic levels.")}
knitr::include_graphics("../output/beta_stack.png")
```

@@

```{r fig5, echo = FALSE,  fig.cap = paste("GDM splines of species-level and interaction-level β-diversity in response to elevational, controlling for geographic proximity and number of sampling dates. The maximum height of each spline represents a given variable’s partial effect on β-diversity (its effect when covariates are held constant), and the shape of each spline represents to the rate of species turnover as it varies along the corresponding gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Dotted vertical lines in the elevation panels indicates the location of the tree line. Within-site comparisons were omitted from GDM analysis to avoid pseudoreplication. Shaded bands depict the uncertainty of each partial effect curve (+/- one standard deviation) based on bootstrapping. Note the differences in scale between the y-axes of each variable.")}
knitr::include_graphics("../output/gdm_stack.png")
```

>>>>

```{r fig1, echo = FALSE,  fig.cap = paste("Study sites in the Berchtesgadener Alps plotted over Google Earth imagery. Sites above the tree line are depecited in red and sites below the tree line in yellow. The elevation of each site is given in meters above sea level.")}
knitr::include_graphics("../output/sites_GE.png")
```

@@

### Elevational distribution of bumble bees and floral resources

We analyzed the abundance of bumble bees and floral resources using separate hierarchical generalized additive models (HGAM) [@Wood2011-pg; @Pedersen2019-kf]. We opted in both models to the collapse temporal variation of abundance within sites and focus on elevational patterns. For bumble bees, we modeled *peak* abundance at each site as a smooth response to elevation, with bumble bee species as a "by" smoothing factor and a random intercept effect. We also included species:year as a random intercept effect to capture annual diffences in species abundance and site as a random intercept effect to avoid overfitting due to pseudoreplication of elevation arising from the measurement of multiple bumble bee species at each site. This construction is a modification of the "type I" model form described in @Pedersen2019-kf. For floral resources, we modeled *mean* flower cover, weighted per bumble be species, as a smooth response to elevation, with bumble bee species and year as "factor-smooths" and site as a random intercept effect (again to avoid overfitting due to pseudoreplication). This construction is a modification of the "type S" model form described in @Pedersen2019-kf.

@@

For our analyses of species and interaction β-diversity, we again opted to focus on elevational patterns by pooling samples within sites, since the sparse observations of individual sampling events tended to inflate β-diversity and its variance to a degree that impaired model fitting and interpretation. After pooling, we calculated the species and interaction β-diversity between all pairs of sites and partitioned total interaction β-diversity into its components of species turnover (changes in species composition) and interaction rewiring (identical species interacting in different ways) [@Novotny2009-my; @Poisot2012-fk]. As an initial analysis, we plotted the relationship between site-wise elevation difference and each metric of β-diversity and verified the significance of the relationship using logistic matrix regression [@Goslee2009-ln], with the difference in number of sampling dates between sites as a covariate to control for potential confounding effects of sampling frequency.

@@

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. GAM analyses were performed with packages `mgcv` [@Wood_2017] and `mgcViz` [@Fasiolo2018aa]. PCoA and species β-diversity calculation were performed with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. Matrix regression was performed with the package `ecodist` [@Goslee2007-aa]. GDM analysis of β-diversity was performed with the package `gdm` [@Fitzpatrick-2021aa]. Annotated `R` code is available in the Supplementary Material.

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species, 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. *B. pascuorum* ("pasc"), *B. pratorum* ("prat"), *B. soroeensis* ("soro"), *B. terrestris/lucorum* ("telu"), and *B. wurflenii* ("wurf") --- accounted for the bulk of overall bumble bee abundance, while *B. hortorum* ("hort"), *B. jonellus* ("jone"), *B. psithyrus* ("psit"), *B. monticola* ("mont"), *B. mendax* ("mend"), *B. mucidus* ("muci"), *B. pyrenaeus* ("pyre"), and *B. gerstaeckeri* ("gers") . *B. hypnorum*, *B. humilis*, and *B. lapidarius* were recorded only sporadically and at low abundance, so we omitted them from all analyses.

@@

```{r fig1, echo = FALSE,  fig.cap = paste("Proportional abundances (cumulative within site) of bumble bee species in our study system, ordered by mean proportional abundance.")}
knitr::include_graphics("../output/bb_abundsum.png")
```

## Elevational distribution of bumble bees and floral resources

Our model explained 71.5% of total deviance (adjusted R^2^ = 0.511).

*B. mendax*, *B. monticola*, *B. mucidus*, and *B. pyrenaeus* exhibited a positive response to elevation and peaked in abundance above the tree line (~ 1500 m.s.l.) (**Figure 3**), consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. The abundance of *B. soroensis* also tended to increase with elevation and peaked above the tree line, but it was also abundant at lower elevations. In contrast, the abundance of *B. pascorum* --- the most abundant species in our study system --- declined steeply with elevation, and a similar pattern was seen for *B. psithyrus*.

while *B. gerstaeckeri* was sparsely distributed at mid-elevation sites.

```{r fig3, echo = FALSE,  fig.cap = paste("Peak bumble bee abundance by elevation and year. Curves shows the conditional GAM predictions excluding the site random effect, and points show original data. Dotted vertical lines indicate the approximate location of the tree line at our study sites. Note different scales of y-axes.")}
knitr::include_graphics("../output/gam_max_bb_abund2.png")
```

```{r fig4, echo = FALSE,  fig.cap = paste("Mean weighted floral abundance (area of flower cover) by elevation and year. Lines shows the conditional GAM predictions excluding the random effects of site and species, and points show original data. The right panel displays the full range of the data while the left panel displays a subset of the range, omitting extreme values of flower cover to highlight the prediction curves.")}
knitr::include_graphics("../output/gam_mean_fl_abund_zoom.png")
```

@@

```{r fig5, echo = FALSE,  fig.cap = paste("Species-level (A) and interaction-level (B, C) plotted against elevational difference. Each point represents the β-diversity (or β-diversity partition) between a pair of sites, and lines represent the overall relationship between β-diversity and elevation difference using binomial regression smooths. Standard errors are not plotted because they would be misleading due to the non-independence inherent to distance matrix regression, but all regressions were significant (p < 0.01). Interaction β-diversity (B, C) is partitioned using Poisot et al.’s (2012) notation: WN = unpartitioned β-diversity, ST = β-diversity due to species turnover, OS =  β-diversity due to interaction rewiring, ST.h =  β-diversity due to species turnover in the higher trophic level (bumble bees), ST.l = β-diversity due to species turnover in the lower trophic level (plants), and ST.lh =  β-diversity due to joint species turnover in higher and lower trophic levels.")}
knitr::include_graphics("../output/beta_stack.png")
```

@@

```{r fig6, echo = FALSE,  fig.cap = paste("GDM splines of species-level and interaction-level β-diversity in response to elevational, controlling for geographic proximity and number of sampling dates. The maximum height of each spline represents a given variable’s partial effect on β-diversity (its effect when covariates are held constant), and the shape of each spline represents to the rate of species turnover as it varies along the corresponding gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Dotted vertical lines in the elevation panels indicates the location of the tree line. Within-site comparisons were omitted from GDM analysis to avoid pseudoreplication. Shaded bands depict the uncertainty of each partial effect curve (+/- one standard deviation) based on bootstrapping. Note the differences in scale between the y-axes of each variable.")}
knitr::include_graphics("../output/gdm_stack.png")
```
